### 2017-10-02 chapters/introduction/patient.tex
Incorporate Sams suggestions

\section{Personalization of Cardiac Biomechanics}
\label{sec:intro_personalization}
In this section we will go through the steps needed to personalize the
mechanical model developed in the previous section. We will start by
giving a brief overview of the interaction between medical imaging and
cardiac computational modeling. Then we will explain how the geometries
are extracted from the medical images and turned into computational
geometries which we can use for finite element computations. Finally we
will explain how to incorporate additional data into model by means of
adjoint based data assimilation techniques.

@@

Since the discovery of X-ray in 1895, medical imaging has played a central
role in diagnosis, treatment planning and follow-up. Without the huge
advancements in medical imaging the last decades, computational models
would not have been as important and promising as it has become.

Cardiac computational modeling often relies on data obtained using
medical imaging techniques, and the quality of such data is therefore of
fundamental importance \cite{lamata2014images}. It is clear, that
without advancements in medical imaging technology, the role of
computational models would be limited to simple, idealized cases. With
access to high quality images, detailed information about cardiac structure
can be used to e.g build anatomically correct models. The
computational models can then later be validated using the same
images, or additional data can be extracted from the images in order to
compute potential biomarkers. Figure \ref{fig:cardiac_imaging_model} summarizes
this interaction between medical images and computational models.

@@

The images are later processed using some image segmentation tool.
One such tool is EchoPac, which is used for analyzing images acquired with
the echo scanner from GE Vingmed. The 4D Auto LVQ tool in EchoPac, is
a tool for processing 3D echo images, and traces of volume,
triangulated surfaces and 3D strain traces \cite{heimdal20114d},
together with a structured mesh of the American Heart Association (AHA) segments
\cite{cerqueira2002standardized} for each time point can be extracted, see Figure
\ref{fig:echopac_output}.

@@

\subsubsection{Mesh generation}
In this section we will explain how to generate a left ventricular
mesh based on segmented surfaces coming from 4D echocardiography.
Figure \ref{fig:echopac_output} shows an example of how the exported data
from the image segmentation tool looks like. For each time point we
can extract triangulated surfaces of the endocardium and epicardium
which can be seen in Figure \ref{fig:echopac_out_surf}. Together with
these surfaces we are also given a so called strain mesh (Figure
\ref{fig:echopac_out_strain_mesh}) located approximately in the
midwall, which defines the approximate location of the AHA-segments
\cite{cerqueira2002standardized}, and can be used to orient the surfaces.

As discussed in Section \ref{sec:mech_boudary}, we constrain the basal movement in
the longitudinal direction, which is easiest to accomplish using a flat
base located at a prescribed location. In order to make the base flat,
we first orient the surfaces so that the longitudinal axis is aligned
with the $x$-axis, and the apex pointing in the positive $x$
direction. To construct the basal plane we first take out the basal
points from the strain mesh, and use these points to construct a least
square fitting plane (Figure \ref{fig:strain_mesh_plane}). Let $(x_i, y_i, z_i), i = 1, \cdots, N$
be the basal points, and suppose the basal plane solves the equation
$z = ax + by + c$, for some unknown constants $a,b,c$. Following a
least square approach, we select the parameters $(a,b,c)$ that
minimizes to sum
\begin{align}
  \sum_{i = 1}^{N} \left(ax_i + by_i + c  - z_i \right)^2.
\end{align}
Once the parameters for the basal plane is found we adjust the size of
the cut, by moving the plane along the longitudinal axis (here
$x-$axis), until the cavity volume agrees with the measured volume
given by the image segmentation tool within some specified tolerance. 
When the correct cut size is found, the points above the plane is removed,
and the remaining surfaces are smoothed using the mesh generation tool GAMer
\cite{yu2008feature}.

@@

\begin{align}
  \begin{split}
    x &= a \sinh \mu \sin \nu \cos \theta,  \\
    y &= a \sinh \mu \sin \nu \sin \theta,  \\
    z &= a \cosh \mu \cos \nu,
  \end{split}
\end{align}
where $a$ is the focal point, $\nu \in [0, \pi]$ is the longitudinal
coordinate, $\theta \in [0, 2\pi]$ is the azimuth angle and $\mu$
is the radial coordinate. For an ellipse given by the equation
$\frac{x^2}{b^2} +\frac{y^2}{c^2}  = 1$ with $b$ being the semi-major
axis, and $c$ the semi-minor axis, the focal point is given by
\begin{align}
  a =\sqrt{b^2 - c^2}.
  \label{eq:focal_point}
\end{align}
The prolate spheroidal coordinate axes are illustrated in
Figure~\ref{fig:prolate_coord}. The inverse relation, going from the
Cartesian coordinate $(x,y,z)$ to the prolate spheroidal coordinate
$(\mu, \nu, \theta)$ given the focal point $a$, is a bit more
intricate, but can be computed as follows 
\begin{align}
  \begin{split}
    \nu &= \arccos (\tau), \\
    \mu &= \arccosh (\sigma), \\
    \theta &= \arctan \frac{z}{y}, \\
    \tau &= \frac{1}{2a} \big( \sqrt{(x + a)^2 + y^2 + z^2} \\
    & \quad + \sqrt{(x - a)^2 + y^2 + z^2} \big), \\
    \sigma &= \frac{1}{2a} \big( \sqrt{(x + a)^2 + y^2 + z^2}\\
    & \quad - \sqrt{(x - a)^2 + y^2 + z^2}\big). 
  \end{split}
\end{align}
It is possible to estimate the focal point by taking $b$ in
\eqref{eq:focal_point} to be the maximum distance from the base to the
apex, and $c$ the maximum radius at the base. Using this estimate,
the base is assumed to be located approximately at the center of the
ellipsoid, i.e $\nu = \pi / 2$ \cite{land2015verification}.
% Compared to geometries used in
% benchmarking of cardiac mechanics problems, this is not a totally
% wrong assumption 
\begin{figure}[htbp]
  \centering
  \begin{subfigure}[t]{0.4\textwidth}
    \includegraphics[width=\textwidth]{chapters/introduction/figures/geometry/prolate.png}
    \caption{\label{fig:prolate_coord}}
  \end{subfigure}
  \begin{subfigure}[t]{0.45\textwidth}
    \includegraphics[width=\textwidth]{chapters/introduction/figures/geometry/prolate_cube.png}
    \caption{\label{fig:prolate_cube}}
  \end{subfigure}
\caption{Illustration of the prolate spheroidal coordinate system.}
\label{fig:prolate}
\end{figure}

This coordinate system can be used when e.g marking the mesh according
to the AHA segments

\subsection{Choosing the reference geometry}
\label{sef:reference_geometry}
% The choice of reference geometry ..
%  choosing the geometry with the smallest
% volume as the reference geometry (i.e the end-systolic geometry)
% (references), the mid-diastolic or the end-diastolic ??
A general problem in biomechanics is that the geometry extracted from
imaging data is not the stress-free geometry, meaning the the geometry
that we observe is subjected to a physical load, e.g blood
pressure. In order for the stress and strain to be correct, we need to
find a geometry, so that when loaded with the measured loads, gives
back the geometry observed in the medical images. 
This is analogous to the problem of finding the geometry of a deflated balloon, given
that we have an image of and inflated version of the balloon. Having
this picture in mind, it is intuitive that such a geometry not necessarily is unique.
Several techniques have been applied in order to estimate the unloaded
geometry, such as the inverse design analysis (ID)
\cite{govindjee1996computational} or the modified updated Lagrangian
formulation (MULF) \cite{gee2010computational}. Other approaches
includes using the mid-diastolic or end-systolic geometry. Another popular
technique is a fixed point iteration scheme,  known as the
backward displacement method \cite{bols2013computational}. Figure
\ref{fig:unloading_lv} illustrate the backward displacement method applied
to an LV geometry. We briefly explain the steps in this methods.

@@

Suppose you are given geometry $\Omega_{\mathcal{I}}$ taken from some
medical image, which is loaded with a pressure $p_{\mathcal{I}}^{\mathrm{LV}}$ (assume LV only of
simplicity). We want to find the unloaded geometry 
$\Omega_{\mathcal{U}}$, so that when we load $\Omega_{\mathcal{U}}$ with $p_{\mathcal{I}}^{\mathrm{LV}}$,
you get back the starting geometry $\Omega_{\mathcal{I}}$. Let
$\Xvec_{\mathcal{I}}$ denote the coordinates in $\Omega_{\mathcal{I}}$, and
let $\mathbf{d}$ denote the displacement field which is the solution
of \eqref{eq:intro_variational_form} after applying the pressure
$p_{\mathcal{I}}^{\mathrm{LV}}$. What we want, is to find
$\Xvec_{\mathcal{U}}$ so that $\Xvec_{\mathcal{I}} =
\Xvec_{\mathcal{U}} + \mathbf{d}(\Xvec_{\mathcal{U}})$. To find
$\Xvec_{\mathcal{U}}$ we let 
\begin{align}
  \Xvec_{n+1} = g(\Xvec_n) =  \Xvec_{\mathcal{I}} - \mathbf{d}(\Xvec_n),
  && \Xvec_0 = \Xvec_{\mathcal{I}}.
\end{align}
According to Banach Fixed Point Theorem, the mapping $g$ has a
fixed-point if $\| \nabla g \|_{\infty} = \| \nabla \mathbf{d}
\|_{\infty} < 1$. This means that as long as the deformation
resulting from the applied pressure, do not result in a stretch that
is more than $100 \%$, this method will converge. Of course the
question about convergence of the fixed point method is one
thing. Another thing is the convergence of the nonlinear solver.
For bi-ventricular geometries, the backward displacement method might
fail to converge, especially if the tissue is soft and the pressure in
the right ventricle is high. In this case the
right ventricular free wall might collided with the septal wall when
subtracting the displacement. An example of this is illustrated in
Figure~\ref{fig:unloading_fail}.

@@

An alternative to the backward displacement method that do not
necessary converge towards a fixed point but provides a more stable
way of obtaining an unloaded configuration was proposed in Ragahavan
et. al \cite{raghavan2006non}. In this algorithm, only one iteration of the
backward displacement method is applied but the subtracted
displacement is multiplied by a scalar $k$, which could be estimated
to minimizing some residual.
% \begin{align}
%   \Xvec_{\mathcal{U}}(k) = \Xvec_{\mathcal{I}} - k \cdot \mathbf{d}(\Xvec_{\mathcal{I}}).
% \end{align}

@@

\subsubsection{Coupled material parameter estimation and unloading
  algorithm}
\label{sec:intro_coupled_material}
One thing to note is that the unloaded configuration depends upon the
chosen material parameters, which are generally also unknown. For
example, applying the backward displacement method to a geometry will
result in a smaller unloaded cavity volume if the material is softer.
Therefore, the material parameters estimation is coupled to the
problem of finding the unloaded, zero pressure geometry.
In Reference \cite{nikou2016effects}, Nikou et al propose a method for estimating
material parameters and the unloaded geometry by iterating between
the two methods, and terminating if the difference in unloaded cavity
volume is less than $5 \%$. This method is general and could be applied
using any unloading algorithm and any method to estimate the material
parameters. In the last paper in this thesis, we analyze this method
and shows that it actually converges to the correct solution are are
stable to noise in the measurements.

@@

\subsection{Data Assimilation}
The constitutive laws in cardiac mechanics comes from experiments
done on tissue slabs, and relates stress in the material to strain.
If we had the same data available for a given patient we could easily
have performed a regression analysis in order to identify the
parameters in the constitutive model. However, this type of data
typically requires that tissue samples are taken from the myocardium,
which is not an option when dealing with living humans. In general,
when dealing with physical systems such as the heart or the weather,
it is not always possible to design an experiment which
allows to determine all properties of a physical system
\cite{chapelle2013fundamental}. In such cases we have to take the
measurement that we have, and do our best to incorporate them into the
model.

@@

The other approach, called variational data assimilation, also known as
4D-var, is based on minimizing a cost functional that
represents the mismatch between observations and simulations, with
possibly additional regularization terms added to the objective
functional. We will focus this approach since this is the approach
taken in this thesis. For a more thorough review of other
data assimilation techniques we refer to the paper by Chapelle
et. al \cite{chapelle2013fundamental}.

@@

\paragraph{The strain observation operator
  $\mathcal{H}_{\mathrm{strain}}$}
Another observation that is encountered in this thesis is strain
measurements, or more precisely average regional strain in the
circumferential, radial or longitudinal direction. Let $\Omega_j$
denote the volume for which the strain should be averaged over, and
let $\mathbf{e}_k$ denote the unit vector field in the preferred strain
direction. Then for a given strain tensor $\mathbf{A}$ we define 
\begin{align}
  \mathcal{H}_{\mathrm{strain}}(\uvec) = \frac{1}{|\Omega_j|}\int_{\Omega_j} \mathbf{e}_k^T \mathbf{A}(\uvec) \mathbf{e}_k  \mathrm{d}V
\end{align}
The strain tensor $ \mathbf{A}$ is typically chosen to be the
Green-Lagrange strain tensor $\mathbf{E}$, or
the material displacement gradient tensor $\F - \I$.

@@

The typical way solving the problem \eqref{eq:pde_constrained} is to
turn the constrained problem into an unconstrained problem. One way
to do this is to consider the \emph{reduced functional}
\begin{align}
  \widehat{\mathcal{J}}(\mu) = \mathcal{J}(\state(\mu), \mu).
\end{align}
Since $\widehat{\mathcal{J}}$ is a pure function of $\mu$, we have
eliminated the constraint and turned the problem into an unconstrained
problem:
\begin{equation}
  \begin{aligned}
    \label{eq:pde_unconstrained}
    & \underset{\mu \in P}{\text{minimize}}
    & &  \widehat{\mathcal{J}}(\mu).
  \end{aligned}
\end{equation}

@@

\subsubsection{Regularization}
\label{sec:intro_regularization}
Regularization is an important concept in many scientific disciplines
concerned with data fitting. When using data that essentially are
realizations of a stochastic process, we need to take into account
that the data are corrupted with noise. Moreover, the model we are using
is a simplification of a physical process, and fitting the
data exactly to the model is unrealistic. This is especially
important if the number of parameters we are trying to estimate are
more than the number of data points used as input. In such cases it is often
beneficial to make some assumptions about the solution we are looking
for in order to avoid overfitting. For example, we may assume that the
parameter we are searching for is smooth. Suppose that $\mu \in
H^1(\Omega)$ is the parameter we want to find, and that we want to
favor smooth solutions. Then the norm of the gradient $\| \nabla \mu
\|$ should be small. One way to achieve this is to add a penalty term
to the objective functional so that in stead of minimizing
$\mathcal{J}(\state, \mu)$ we minimize $\mathcal{J}(\state, \mu) +
\lambda \| \nabla \mu\|$, where the constant $\lambda$, known as the
regularization parameter, controls the smoothness of $\mu$.
Note that in the limit $\lambda \rightarrow \infty$ we would end up
with a constant value of $\mu$. This approach is referred to as Tikhonov
regularization.

@@

Another problem, which is also related to identifiability is
the question about uniqueness of the solution. A practical test is to
start the optimization algorithm from different initial points and see
if you end up at the same optimum. If this does not happen, it is
likely that the objective functional is not convex, which is a
requirement for uniqueness. Unfortunately, this is the case for many
problems in PDE-constrained optimization. In these cases, the objective
functional might have several distinct local minima, in which case
different techniques for finding the best local minima exists
\cite{farrell2015multiple}.

@@

As can be seen from Figure \ref{fig:cardiac_imaging_model}, data
assimilation of computational models and medical imaging can be used to
extract biomarkers. By biomarker, we mean an indicator of some
biological of physiological process that give insight into the state
or condition of that process. Examples of biomarkers that can be
extracted from a computational model are parameters related to tissue
stiffness, like the parameters in the Holzapfel-Ogden material law
\eqref{eq:holzapel_trans}, parameters related to contractility like
the active stress ($T_a$ in \eqref{eq:intro_active_stress}) or the
active strain ($\gamma$ in \eqref{eq:intro_active_strain_Fa_gjerald}),
or estimates or myofiber stress.
Computational models opens a new door when it comes to extraction of
biomarkers, and mechanical features such as stress, contractility and
elastance can easily be computed using a computational model. Of
course, validation of these models is what matter most in order to
make these biomarkers clinically useful \cite{caruel2014dimensional}.

>>>>

\section{Personalization of Cardiac Biomechanics}
\label{sec:intro_personalization}
In this section we will go through the steps needed to personalize the
mechanical model developed in the previous section. We will start by
giving a brief overview of the interaction between medical imaging and
cardiac computational modeling. Then we will explain how the geometries
are extracted from the medical images and turned into computational
geometries that we can use for finite element computations. Finally we
will explain how to incorporate additional data into model by means of
adjoint based data assimilation techniques.

@@

Since the discovery of X-ray in 1895, medical imaging has played a central
role in diagnosis, treatment planning and follow-up. Without the huge
advancements in medical imaging the last decades, computational models
would not have been as important and promising as they have become.

Cardiac computational modeling often relies on data obtained using
medical imaging techniques, and the quality of such data is therefore of
fundamental importance \cite{lamata2014images}. It is clear that
without advancements in medical imaging technology, the role of
computational models would be limited to simple, idealized cases. With
access to high quality images, detailed information about cardiac structure
can be used to e.g build anatomically correct models. The
computational models can then later be validated using the same
images, or additional data can be extracted from the images in order to
compute potential biomarkers. Figure \ref{fig:cardiac_imaging_model} summarizes
this interaction between medical images and computational models.

@@

The images are further processed using some image segmentation tool.
One such tool is EchoPac, which is used for analyzing images acquired with
the echo scanner from GE Vingmed (Horten, Norway). The 4D Auto LVQ tool in EchoPac, is
a tool for processing 3D echo images, and can be used to extract traces of volume,
triangulated surfaces and 3D strain traces \cite{heimdal20114d},
together with a structured mesh of the American Heart Association (AHA) segments
\cite{cerqueira2002standardized} for each time point, see Figure
\ref{fig:echopac_output}.

@@

\subsubsection{Mesh generation}
In this section we will explain how to generate a left ventricular
mesh based on segmented surfaces coming from 4D echocardiography.
Figure \ref{fig:echopac_output} shows an example of how the exported data
from the image segmentation tool look like. For each time point we
can extract triangulated surfaces of the endocardium and epicardium
which can be seen in Figure \ref{fig:echopac_out_surf}. Together with
these surfaces we are also given a so called strain mesh (Figure
\ref{fig:echopac_out_strain_mesh}) located approximately in the
midwall, which defines the approximate location of the AHA-segments
\cite{cerqueira2002standardized}, and can be used to orient the surfaces.

As discussed in Section \ref{sec:mech_boudary}, we constrain the basal movement in
the longitudinal direction, which is easiest to accomplish using a flat
base located at a prescribed location. In order to make the base flat,
we first orient the surfaces so that the longitudinal axis is aligned
with the $x$-axis, and the apex pointing in the positive $x$
direction. To construct the basal plane we first take out the basal
points from the strain mesh, and use these points to construct a least
square fitting plane (Figure \ref{fig:strain_mesh_plane}). Let $(x_i, y_i, z_i), i = 1, \cdots, N$
be the basal points, and suppose the basal plane solves the equation
$z = ax + by + c$, for some unknown constants $a,b,c$. Following a
least square approach, we select the parameters $(a,b,c)$ that
minimizes to sum
\begin{align}
  \sum_{i = 1}^{N} \left(ax_i + by_i + c  - z_i \right)^2.
\end{align}
Once the parameters for the basal plane are found we adjust the size of
the cut, by moving the plane along the longitudinal axis (here
$x-$axis), until the cavity volume agrees with the measured volume
given by the image segmentation tool within some specified tolerance. 
When the correct cut size is found, the points above the plane is removed,
and the remaining surfaces are smoothed using the mesh generation tool GAMer
\cite{yu2008feature}.

@@

\begin{align}
  \begin{split}
    x &= a \sinh \mu \sin \nu \cos \theta,  \\
    y &= a \sinh \mu \sin \nu \sin \theta,  \\
    z &= a \cosh \mu \cos \nu,
  \end{split}
\end{align}
where $a$ is the focal point, $\nu \in [0, \pi]$ is the \emph{longitudinal}
coordinate, $\theta \in [0, 2\pi]$ is the \emph{circumferential} azimuth angle and $\mu$
is the \emph{radial} coordinate. For an ellipse given by the equation
$\frac{x^2}{b^2} +\frac{y^2}{c^2}  = 1$ with $b$ being the semi-major
axis, and $c$ the semi-minor axis, the focal point is given by
\begin{align}
  a =\sqrt{b^2 - c^2}.
  \label{eq:focal_point}
\end{align}
The prolate spheroidal coordinate axes are illustrated in
Figure~\ref{fig:prolate_coord}. The inverse relation, going from the
Cartesian coordinate $(x,y,z)$ to the prolate spheroidal coordinate
$(\mu, \nu, \theta)$ given the focal point $a$, is a bit more
intricate, but can be computed as follows 
\begin{align}
  \begin{split}
    \nu &= \arccos (\tau), \\
    \mu &= \arccosh (\rho), \\
    \theta &= \arctan \frac{z}{y}, \\
    \tau &= \frac{1}{2a} \big( \sqrt{(x + a)^2 + y^2 + z^2} \\
    & \quad + \sqrt{(x - a)^2 + y^2 + z^2} \big), \\
    \rho &= \frac{1}{2a} \big( \sqrt{(x + a)^2 + y^2 + z^2}\\
    & \quad - \sqrt{(x - a)^2 + y^2 + z^2}\big). 
  \end{split}
\end{align}
It is possible to estimate the focal point by taking $b$ in
\eqref{eq:focal_point} to be the maximum distance from the base to the
apex, and $c$ the maximum radius at the base. Using this estimate,
the base is assumed to be located approximately at the center of the
ellipsoid, i.e $\nu = \pi / 2$ \cite{land2015verification}.
% Compared to geometries used in
% benchmarking of cardiac mechanics problems, this is not a totally
% wrong assumption 
\begin{figure}[htbp]
  \centering
  \begin{subfigure}[t]{0.4\textwidth}
    \includegraphics[width=\textwidth]{chapters/introduction/figures/geometry/prolate.png}
    \caption{\label{fig:prolate_coord}}
  \end{subfigure}
  \begin{subfigure}[t]{0.45\textwidth}
    \includegraphics[width=\textwidth]{chapters/introduction/figures/geometry/prolate_cube.png}
    \caption{\label{fig:prolate_cube}}
  \end{subfigure}
\caption{Illustration of the prolate spheroidal coordinate system.}
\label{fig:prolate}
\end{figure}

This coordinate system can be used when e.g marking the mesh according
to the AHA segments and to create vector fields in the longitudinal,
circumferential and radial direction. These vector fields are useful
when comparing the simulated results to the image-based measurements.

\subsection{Choosing the reference geometry}
\label{sef:reference_geometry}
% The choice of reference geometry ..
%  choosing the geometry with the smallest
% volume as the reference geometry (i.e the end-systolic geometry)
% (references), the mid-diastolic or the end-diastolic ??
A general problem in biomechanics is that the geometry extracted from
imaging data is not the stress-free geometry, meaning the the geometry
that we observe is subjected to a physical load, e.g blood
pressure. In order for the stress and strain to be correct, we need to
find a geometry, so that when loaded with the measured loads, gives
back the geometry observed in the medical images. 
This is analogous to the problem of finding the geometry of a deflated balloon, given
that we have an image of an inflated version of the balloon. Having
this picture in mind, it is intuitive that such a geometry not necessarily is unique.
Several techniques have been applied in order to estimate the unloaded
geometry, such as the inverse design analysis (ID)
\cite{govindjee1996computational} or the modified updated Lagrangian
formulation (MULF) \cite{gee2010computational}. Other approaches
includes using the mid-diastolic or end-systolic geometry. Another popular
technique is a fixed point iteration scheme,  known as the
backward displacement method \cite{bols2013computational}. Figure
\ref{fig:unloading_lv} illustrate the backward displacement method applied
to an LV geometry. We briefly explain the steps in this methods.

@@

Suppose you are given geometry $\Omega_{\mathcal{I}}$ taken from some
medical image, which is loaded with a pressure $p_{\mathcal{I}}^{\mathrm{LV}}$ (assume LV only of
simplicity). We want to find the unloaded geometry 
$\Omega_{\mathcal{U}}$, so that when we load $\Omega_{\mathcal{U}}$ with $p_{\mathcal{I}}^{\mathrm{LV}}$,
you get back the starting geometry $\Omega_{\mathcal{I}}$. Let
$\Xvec_{\mathcal{I}}$ denote the coordinates in $\Omega_{\mathcal{I}}$, and
let $\mathbf{d}$ denote the displacement field which is the solution
of \eqref{eq:intro_variational_form} after applying the pressure
$p_{\mathcal{I}}^{\mathrm{LV}}$. What we want, is to find
$\Xvec_{\mathcal{U}}$ so that $\Xvec_{\mathcal{I}} =
\Xvec_{\mathcal{U}} + \mathbf{d}(\Xvec_{\mathcal{U}})$. To find
$\Xvec_{\mathcal{U}}$ we let 
\begin{align}
  \Xvec_{n+1} = g(\Xvec_n) =  \Xvec_{\mathcal{I}} - \mathbf{d}(\Xvec_n),
  && \Xvec_0 = \Xvec_{\mathcal{I}}.
\end{align}
According to Banach Fixed Point Theorem, the mapping $g$ has a
fixed-point if $\| \nabla g \|_{\infty} = \| \nabla \mathbf{d}
\|_{\infty} < 1$. This means that as long as the deformation
resulting from the applied pressure does not result in a stretch that
is more than $100 \%$, this method will converge. Of course the
question about convergence of the fixed point method is one
thing. Another thing is the convergence of the nonlinear solver.
For bi-ventricular geometries, the backward displacement method might
fail to converge, especially if the tissue is soft and the pressure in
the right ventricle is high. In this case the
right ventricular free wall might collided with the septal wall when
subtracting the displacement. An example of this is illustrated in
Figure~\ref{fig:unloading_fail}.

@@

An alternative to the backward displacement method that does not
necessary converge towards a fixed point but provides a more stable
way of obtaining an unloaded configuration was proposed in Ragahavan
et. al \cite{raghavan2006non}. In this algorithm, only one iteration of the
backward displacement method is applied but the subtracted
displacement is multiplied by a scalar $k$, which could be estimated
to minimizing some residual.
% \begin{align}
%   \Xvec_{\mathcal{U}}(k) = \Xvec_{\mathcal{I}} - k \cdot \mathbf{d}(\Xvec_{\mathcal{I}}).
% \end{align}

@@

\subsubsection{Coupled material parameter estimation and unloading
  algorithm}
\label{sec:intro_coupled_material}
One thing to note is that the unloaded configuration depends upon the
chosen material parameters, which are generally also unknown. For
example, applying the backward displacement method to a geometry will
result in a smaller unloaded cavity volume if the material is softer.
Therefore, the material parameters estimation is coupled to the
problem of finding the unloaded, zero pressure geometry.
In \cite{nikou2016effects}, Nikou et al proposed a method for estimating
material parameters and the unloaded geometry by iterating between
the two methods, and terminating if the difference in unloaded cavity
volume is less than $5 \%$. This method is general and could be applied
using any unloading algorithm and any method to estimate the material
parameters. In the last paper in this thesis, we analyze this method
and show that it actually converges to the correct solution and is
stable to noise in the measurements.

@@

\subsection{Data Assimilation}
The constitutive laws in cardiac mechanics comes from experiments
done on tissue slabs, and relates stress in the material to strain.
If we had the same data available for a given patient we could easily
have performed a regression analysis to identify the
parameters in the constitutive model. However, this type of data
typically requires that tissue samples are taken from the myocardium,
which is not an option when dealing with living humans. In general,
when dealing with physical systems such as the heart or the weather,
it is not always possible to design an experiment which
allow us to determine all properties of a physical system
\cite{chapelle2013fundamental}. In such cases we have to take the
measurements that we have, and do our best to incorporate them into the
model.

@@

The other approach which is taken in this thesis is called variational
data assimilation, and also referred as 4D-var. This approach is based
on minimizing a cost functional that represents the mismatch between
observations and simulations, with possibly additional regularization
terms added to the objective functional.
For a more thorough review of other
data assimilation techniques we refer to the paper by Chapelle
et. al \cite{chapelle2013fundamental}.

@@

\paragraph{The strain observation operator
  $\mathcal{H}_{\mathrm{strain}}$}
Another observation that is encountered in this thesis are strain
measurements, or more precisely average regional strain in the
circumferential, radial or longitudinal direction. Let $\Omega_j$
denote the volume for which the strain should be averaged over, and
let $\mathbf{e}_k$ denote the unit vector field in the preferred strain
direction. Then for a given strain tensor $\mathbf{A}$ we define 
\begin{align}
  \mathcal{H}_{\mathrm{strain}}(\uvec) = \frac{1}{|\Omega_j|}\int_{\Omega_j} \mathbf{e}_k^T \mathbf{A}(\uvec) \mathbf{e}_k  \mathrm{d}V
\end{align}
The strain tensor $ \mathbf{A}$ is typically chosen to be the
Green-Lagrange strain tensor $\mathbf{E}$, or
the material displacement gradient tensor $\F - \I$.

@@

The typical way of solving the problem \eqref{eq:pde_constrained} is to
turn the constrained problem into an unconstrained problem. One way
to do this is to consider the \emph{reduced functional}
\begin{align}
  \widehat{\mathcal{J}}(\mu) = \mathcal{J}(\state(\mu), \mu).
\end{align}
Since $\widehat{\mathcal{J}}$ is a pure function of $\mu$, we have
eliminated the constraint and turned the problem into an unconstrained
problem:
\begin{equation}
  \begin{aligned}
    \label{eq:pde_unconstrained}
    & \underset{\mu \in P}{\text{minimize}}
    & &  \widehat{\mathcal{J}}(\mu).
  \end{aligned}
\end{equation}

@@

\subsubsection{Regularization}
\label{sec:intro_regularization}
Regularization is an important concept in many scientific disciplines
concerned with data fitting. When using data that essentially are
realizations of a stochastic process, we need to take into account
that the data are corrupted with noise. Moreover, the model we are using
is a simplification of a physical process, and fitting the
data exactly to the model is unrealistic. This is especially
important if the number of parameters we are trying to estimate are
more than the number of data points used as input. In such cases it is often
beneficial to make some assumptions about the solution we are looking
for in order to avoid overfitting. For example, we may assume that the
parameter we are searching for is smooth. Suppose that $\mu \in
H^1(\Omega)$ is the parameter we want to find, and that we want to
favor smooth solutions. Then the norm of the gradient $\| \nabla \mu
\|$ should be small. One way to achieve this is to add a penalty term
to the objective functional so that instead of minimizing
$\mathcal{J}(\state, \mu)$ we minimize $\mathcal{J}(\state, \mu) +
\lambda \| \nabla \mu\|$, where the constant $\lambda$, known as the
regularization parameter, controls the smoothness of $\mu$.
Note that in the limit $\lambda \rightarrow \infty$ we would end up
with a constant value of $\mu$. This approach is referred to as Tikhonov
regularization.

@@

Another problem, which is also related to identifiability is
the question about uniqueness of the solution. A practical test is to
start the optimization algorithm from different initial points and see
if you end up at the same optimum. If this does not happen, it is
likely that the objective functional is not convex, which is a
requirement for uniqueness. Unfortunately, this is the case for many
problems in PDE-constrained optimization. In these cases, the objective
functional might have several distinct local minima, in which case
different techniques for finding the best local minima exist
\cite{farrell2015multiple}.

@@

As can be seen from Figure \ref{fig:cardiac_imaging_model}, data
assimilation of computational models and medical imaging can be used to
extract biomarkers. By biomarker, we mean an indicator of some
biological of physiological process that gives insight into the state
or condition of that process. Examples of biomarkers that can be
extracted from a computational model are parameters related to tissue
stiffness, like the parameters in the Holzapfel-Ogden material law
\eqref{eq:holzapel_trans}, parameters related to contractility like
the active stress ($T_a$ in \eqref{eq:intro_active_stress}) or the
active strain ($\gamma$ in \eqref{eq:intro_active_strain_Fa_gjerald}),
or estimates or myofiber stress.
Computational models open a new door when it comes to extraction of
biomarkers, and mechanical features such as stress, contractility and
elastance can easily be computed using a computational model. Of
course, validation of these models is what matter most in order to
make these biomarkers clinically useful \cite{caruel2014dimensional}.
